---
title: "Class 12 Homework"
author: "Saket Chodavarapu (PID: A18582086)"
format: pdf
toc: TRUE
---

## Q13

Read in the file.

```{r}
data <- read.table("https://bioboot.github.io/bimm143_W26/class-material/rs8067378_ENSG00000172057.6.txt")

head(data)
summary(data)
table(data$geno)
```

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

The sample size for A/A is 108, A/G is 233, and G/G is 121.

```{r}
library(dplyr)

data_aa <- data %>% filter(geno=="A/A")
summary(data_aa)
```

```{r}
data_ag <- data %>% filter(geno=="A/G")
summary(data_ag)
```

```{r}
data_gg <- data %>% filter(geno=="G/G")
summary(data_gg)
```

The median expression levels for A/A is 31.25, A/G is 25.065, and G/G is 20.074.

## Q14: Boxplot

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(data, aes(x=geno, y=exp)) +
  geom_boxplot()
```

A/A seems to have a higher level of expression than G/G. The SNP does affect the expression of ORMDL3, based on visual observation of the boxplot. A/A has a higher min, Q1, median, Q3, and max than those of G/G. Also, A/A's Q1 is greater than G/G's Q3.





















