---
title: "Class 5: Data Viz with ggplot"
author: "Saket Chodavarapu (PID: A18582086)"
format: pdf
toc: true
---

## Background

There are lots of ways to make plots in R. These include so-called "base R" (like the `plot()`) and add-on packages like **ggplot2**.

Let's make the same plot with these two graphics systems. We can use the inbuilt `cars` dataset:

```{r}
head(cars)
```

With "base R" we can simply: 

```{r}
plot(cars)
```

Now let's try ggplot. First I need to install the package using `install.packages("ggplot2")`.

> **Note:** We never run an `install.packages()` in a code chunk otherwise it will re-install needlessly every time we render our document.

Every time we want to use an add-on package we need to load it up with a call to `library()`

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot needs at least 3 things:

1. The **data** i.e. stuff to plot as a data.frame
2. The **aes** or aesthetics that map the data to the plot
3. The **geom_** or geometry i.e. the plot type such as points, lines, etc.

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_line()
```

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(x="Speed (mph)",
       y="Distance (ft)",
       title="Stopping Distance of Old Cars") +
  theme_bw()
```

## Gene Expression Plot

Read some data on the effects of GLP-1 inhibitor (drug) on gene expression values:

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Version 1 plot - start simple by getting some ink on the page.

```{r}
ggplot(genes) +
  aes(Condition1, Condition2) +
  geom_point(col="blue", alpha=0.2)
```

Let's color by `State` up, down, or no change.

```{r}
table(genes$State)
```

```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point() +
  scale_color_manual(values = c("blue",
                                "gray",
                                "red")) +
  labs(x = "Control (no drug)",
       y = "Drug Treatment",
       title = "Gene Expression Changes with GLP-1 Drug") +
  theme_minimal()
```

## Going further with gapminder

Here we explore the famous `gapminder` dataset with some custom plots.

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many rows does this dataset have?

```{r}
nrow(gapminder)
```

> How many different continents are in this dataset?

```{r}
table(gapminder$continent)
```

Version 1 plot GDP vs LifeExp for all rows

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point()
```

I want to see a plot for each continent - in ggplot lingo this is called "faceting"

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point() +
  facet_wrap(~continent)
```

## First look at the dplyr package

Another add-on package with a function called `filter()` that we want to use.

```{r}
library(dplyr)
```

```{r}
filter(gapminder, year==2007, country=="Ireland")
```

```{r}
input <- filter(gapminder, year == 2007 | year == 1997)

ggplot(input) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point() +
  facet_wrap(~year)
```



