---
title: "Class 18: Investigating Pertussis Resurgence"
author: "Saket Chodavarapu (PID: A18582086)"
format: pdf
toc: TRUE
---

## Investigating pertussis cases by year

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called `cdc` and use **ggplot** to make a plot of cases numbers over time.

```{r}
library(datapasta)

cdc <- data.table::data.table(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L,2020L,2021L,2022L,2023L),
  No_Reported_Pertussis_Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617,6124,
                                   2116,3044,7063)
)

head(cdc)
```

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(Year, No_Reported_Pertussis_Cases) +
  geom_point() +
  geom_line() +
  labs(title="Pertussis Cases by Year (1922-2023)", y="Number of cases")
```

## A tale of two vaccines (wP & aP)

> Q2. Using the ggplot `geom_vline()` function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) +
  aes(Year, No_Reported_Pertussis_Cases) +
  geom_point() +
  geom_line() +
  labs(title="Pertussis Cases by Year (1922-2023)", y="Number of cases") +
  geom_vline(xintercept=1946, col="blue") +
  geom_vline(xintercept=1996, col="red")
```

After the 1946 introduction of the wP vaccine, the number of cases drastically decreased. After the 1996 switch to the aP vaccine, there was an increase in cases but the number started to decrease again.

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine, the number of cases began to rise again, but then dropped back down.
A possible explanation is that people were hesitant to take the newer vaccine.

## Exploring CMI-PB data

```{r}
library(jsonlite)
```

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 

head(subject, 3)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

There are 87 aP and 85 wP infancy vaccinated subjects in the dataset.

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

There are 60 male and 112 female subjects in the dataset.

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

The breakdown of race and biological sex is listed in the table.

### Side-Note: Working with dates

```{r}
library(lubridate)
```

```{r}
today()
today() - ymd("2000-01-01")
time_length( today() - ymd("2000-01-01"),  "years")
```

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
subject$age <- today() - ymd(subject$year_of_birth)

library(dplyr)

wP <- subject %>% filter(infancy_vac == "wP")

round( summary( time_length( wP$age, "years" ) ) )
```

The average age of wP individuals is 37 years.

```{r}
aP <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( aP$age, "years" ) ) )
```

The average age of aP individuals is 28 years.

```{r}
t.test(time_length(wP$age, "years"), time_length(aP$age, "years"))
```

The p-value for the t-test is less than 2.2e-16, which means there is a significant statistical difference between the average age of wP individuals and aP individuals.

> Q8. Determine the age of all individuals at time of boost?

```{r}
days <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(days, "year")
head(age_at_boost)
```

The age of all individuals at time of boost is stored in the variable age_at_boost.

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

Based on the two histograms, the two groups appear to be significantly different.

### Joining multiple tables

```{r}
# Complete the API URLs...
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/plasma_ab_titer", simplifyVector = TRUE)
```

> Q9. Complete the code to join `specimen` and `subject` tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- full_join(specimen, subject)
dim(meta)
head(meta)
```

> Q10. Now using the same procedure join `meta` with `titer` data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```

> Q11. How many specimens (i.e. entries in `abdata`) do we have for each `isotype`?

```{r}
table(abdata$isotype)
```

> Q12. What are the different `$dataset` values in `abdata` and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

The number of rows for the most recent dataset is lower than all the other datasets.

## Examine IgG Ab titer levels

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The antigens PT, PRN, FIM2/3, and FHA show differences in the level of IgG antibody titers recognizing them over time. These antigens show differences because they trigger a secondary immune response, whereas other antigens do not.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```

```{r}
igg %>% filter(visit != 8) %>%
ggplot() +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  xlim(0,75) +
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```

> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw() +
  labs(title="OVA antigen levels per visit (aP red, wP teal)")
```

```{r}
filter(igg, antigen=="PT") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw() +
  labs(title="PT antigen levels per visit (aP red, wP teal)")
```

> Q16. What do you notice about these two antigens time courses and the PT data in particular?

OVA seems to steadily decrease over time for both aP and wP. PT initially increases until visit 5, and then decreases for both aP and wP. OVA levels are generally higher than PT levels for both aP and wP patients.

> Q17. Do you see any clear difference in aP vs. wP responses?

For both of these antigens, there is no clear difference between aP and wP responses.

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

> Q18. Does this trend look similar for the 2020 dataset?

```{r}
abdata.20 <- abdata %>% filter(dataset == "2020_dataset")

abdata.20 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
    coord_cartesian(xlim = c(0, 125)) +
  labs(title="2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

The 2021 trend looks fairly similar to the 2020 dataset. Both show an increase from day 0 to day 14, and then a slight decrease until day 120.

## Obtaining CMI-PB RNASeq data

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
#meta <- inner_join(specimen, subject)
ssrna <- inner_join(rna, meta)
```

> Q19. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of `visit` vs. `tpm`).

```{r}
ggplot(ssrna) +
  aes(visit, tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```

> Q20.: What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

Typically, the maximum expression of this gene occurs during visit 4.

> Q21. Does this pattern in time match the trend of antibody titer data? If not, why not?

The pattern almost matches the trend of antibody titer data, except antigen levels peak around visit 5 while antibody levels peak around visit 4. As the cell produces more antibodies, the antigen levels decrease, which is what we observe in the antigen level plots.

```{r}
ggplot(ssrna) +
  aes(tpm, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```

```{r}
ssrna %>%  
  filter(visit==4) %>% 
  ggplot() +
    aes(tpm, col=infancy_vac) + geom_density() + 
    geom_rug() 
```




























































